Exploring the Design Space of Differentially Private Bayesian Inference under Hellinger Distance

### 2019-01-22 ICML2019/main.tex

\section{Private Mechanisms}
\label{sec_mechs}
\subsection{Calibrating noise w.r.t. $\ell_1$ norm ($\lapmech$)}
We start off by presenting a mechanism based on the Laplace primitive.
The mechanism, denoted by $\lapmech$, takes in input the data $\dataobs$, a mechanism, $\bysinfer$, that is used to perform inference on some parameter,
a prior and a likelihood function, and the $\epsilon$ parameter measuring the privacy level.
Now, $\lapmech(\dataobs,\bysinfer, \mathsf{prior}, \mathsf{likelihood}, \epsilon)$,
simply adds Laplace noise to the parameters of the posterior distributions. The mechanisms also performs a
post-processing phase by flooring and  truncating the noisy parameters so that they belong to the right interval.
In the case of the Beta-Binomial model the output produced is: $\betad(\alpha +  \lfloor{\Delta \alpha + \eta}\rfloor^n_0,\beta + n - \lfloor{\Delta \alpha + \eta}\rfloor^n_0)$
with $\eta \thicksim \lap{0}{\frac{\Delta \bysinfer}{\epsilon}}$, where $\lfloor \cdot \rfloor^n_0$ is taking the floor value and truncating into
$[0,n]$. This mechanism is described by the pesudocode in Figure \ref{mech:lapmech}.\\
Since we are adding noise directly to the parameters, the sensitivity should be calibrated w.r.t $\ell_1$-norm. That is,
the sensitivity is defined as:
\[
   \Delta \bysinfer =\max\limits_{\dataobs, \dataobs' \in \{0,1\}^{n}, ||\dataobs - \dataobs'||_1 \leq 1} ||\bysinfer(\dataobs) - \bysinfer(\dataobs')||_1
\]
This value can be quickly seen as being upper bounded by  the dimensionality. (e.g. 2 for the Beta-Binomial model and  $k$ for
the Dirichlet-Multinomial model).
  \begin{algorithm}
  \caption{$\lapmech$ for the Beta-Binomial model}
  \label{mech:lapmech}
  \begin{algorithmic}
  \INPUT observed data set $\dataobs \in \{0,1\}^n$, inference algorithm: $\bysinfer$, 
  prior: $\betad(\alpha, \beta)$, likelihood: $\binomial{n}{\theta}$, $\epsilon$
  \STATE \quad {\bf let} $\betad(\alpha', \beta') = \bysinfer(\binomial{n}{\theta}, \betad(\alpha, \beta), \dataobs)$.  
  \STATE \quad  $\eta \sim \lap{0}{\frac{2}{\epsilon}}$.
  \STATE \quad  $\tilde{\alpha}=\alpha + \lfloor{(\alpha' - \alpha) + \eta}\rfloor^n_0$ 
  \STATE \quad  $\tilde{\beta}=\beta + n - \lfloor{(\alpha' - \alpha) + \eta}\rfloor^n_0$ \\
  % \RETURN $a_{k+1}$.
  \noindent {\bf return} $\betad(\tilde{\alpha}, \tilde{\beta})$
  \end{algorithmic}
  \end{algorithm}

  In Dirichlet-Multinomial model, $\lapmech$ produces output:
  $\dirichlet{\alpha_1 +  \lfloor{\Delta \alpha_1 + \eta_1}\rfloor^n_0, \cdots, \alpha_k + \lfloor n - \sum_{i = 1}^{k-1}\lfloor{\Delta \alpha_i + \eta_i}\rfloor^n_0 \rfloor^n_0},$
  where $\eta_i \thicksim \lap{0}{\frac{k - 1}{\epsilon}}$.
  We will use $ \lapmech(\dataobs) = \betad(\tilde{\alpha}, \tilde{\beta})$ or $\dirichlet{\tilde{\valpha}}$ for short.

\subsection{Improving the sensitivity ($\ilapmech$)}
 The accuracy of $\lapmech$ can be reduced to a constant.
 Indeed, two adjacent data sets $\dataobs, \dataobs'$  induce posterior distributions -- $\bysinfer(\dataobs), \bysinfer(\dataobs')$ --
 whose parameters differ by at most 2 w.r.t $\ell_1$ norm. This because we can see the parameters of the distributions as an histogram.
 Hence, it is enough to use sensitivity $1$ in 2 dimensions and $2$ in higher dimensions.
 {\color{red} Notice that we are using the bounded notion of differential privacy \cite{????}.}
 \todo{why 1 in 2 dimensions?}
 We denote the mechanism which calibrates the noise to this smaller sensitivity as $\ilapmech$.
 $\ilapmech(\dataobs,\bysinfer, \mathsf{prior}, \mathsf{likelihood}, \epsilon)$ produces:
 $\betad(\alpha +  \lfloor{\Delta \alpha + \eta}\rfloor^n_0,\beta + n - \lfloor{\Delta \alpha + \eta}\rfloor^n_0) $
 where $\eta \thicksim \lap{0}{\frac{1}{\epsilon}}$ in the Beta-binomial model and
 $\dirichlet{\alpha_1 +  \lfloor{\Delta \alpha_1 + \eta_1}\rfloor^n_0, \cdots,  \alpha_k + \lfloor n - \sum_{i = 1}^{k-1}\lfloor{\Delta \alpha_i + \eta_i}\rfloor^n_0 \rfloor^n_0}$
 with $\eta_i \thicksim \lap{0}{\frac{2}{\epsilon}}$ in Dirichlet-multinomial model. Details are shown in Figure \ref{mech:ilapmech}. We will also from now on
 write $\ilapmech(\dataobs)$ when the other parameters are obvious from the context.
  \begin{algorithm}
  \caption{$\ilapmech$ in Dirichlet-Multinomial model}
  \label{mech:ilapmech}
  \begin{algorithmic}
  \INPUT observed data set $\dataobs \in \mathcal{X}^n$, inference algorithm: $\bysinfer$, 
  prior: $\dirichlet{\valpha}$, likelihood: $\multinomial{n}{\vtheta}$, $\epsilon$
  \STATE \quad {\bf let} $\dirichlet{\valpha'} = \bysinfer(\multinomial{n}{\vtheta}, \dirichlet{\valpha}, \dataobs)$, $k = |\mathcal{X}|$.
  \STATE \quad {\bf Initialize} a vector $\tilde{\valpha} = (0, \dots, 0 )$ of $k$ dimensions .
  \STATE \quad {\bf For} $i = 1 \dots (k - 1)$:
  \STATE \quad \quad  {\bf let} $\eta \sim \lap{0}{\frac{2}{\epsilon}}$.
  \STATE \quad \quad  $\tilde{\alpha_i}=\alpha_i + \lfloor{(\alpha_i' - \alpha_i) + \eta}\rfloor^n_0$ 
  % \RETURN $a_{k+1}$.
  \STATE \quad $\tilde{\alpha_k} = \alpha_k + \lfloor n - \sum_{i = 1}^{k-1}\lfloor{(\alpha_i' - \alpha_i) + \eta_i}\rfloor^n_0 \rfloor^n_0$
  \STATE {\bf return} $\dirichlet{\tilde{\valpha}}$
  \end{algorithmic}
  \end{algorithm}

\subsection{A new family of mechanisms}
In this section, we explore  mechanism which we denote by $\expmech$. $\expmech$ is an instance
of the exponential mechanism \cite{??} where the scoring function is based on a metric over
probability distributoins.
The idea is that $\expmech$ will sample and output a distribution with higher probability when
it is closer to the real posterior; importantly,  \emph{closer} is measured using a metric over probability
distributions as opposed to a metric over the parameters that represent it. In this work we will
use the Hellinger distance, but this approach could be extended to other $f$-divergences. We will
start by describing how the mechanism works in the case of the Beta-Binomial model noticing that
the extension to the Dirichelet-Multinomial model is straightforward.
Given a prior distribution $\betad(\alpha, \beta)$ and an observed data set: $\dataobs\in\{0,1\}^n$,
we define the follwing set  where the mechanism will sample from:
\[
  \candidateset=\{\bysinfer(\binomial{n}{\theta}, \betad(\alpha, \beta),\dataobs)\mid \dataobs \in \{0,1\}^n\}.
\]
Where $\bysinfer(\binomial{n}{\theta}, \bprior,\dataobs)$ denotes the posterior beta distribution inferred starting from the parameters in input.
Notice that the above set reduces to:
\[
  \{\betad(\alpha',\beta')\mid \alpha'=\alpha + \Delta\alpha, \beta'=\beta + n-\Delta\alpha\}
\]
We use $\bysinfer(\dataobs)$ to denote the Bayesian inference process on the observed data $\dataobs$
when the other two parameters, that is prior and likelihood, are fixed.

\subsubsection{A first approach: the $\expmech$ mechanism}
$\expmech(x,u,\candidateset)$  outputs an element $r\in\candidateset$ with  probability: % proportional to $\exp(\frac{\epsilon u(x,r)}{2 GS})$.
\[
%\underset{z \thicksim \expmech(x,u,\candidateset) }{\Pr}[z=r] = \frac
\frac
  {\exp(\frac{\epsilon\cdot \ux{r}}{2 GS})}
{\Sigma_{r' \in \mathcal{R}}\exp(\frac{\epsilon\cdot \ux{r'}}{2\cdot GS})},
\]
where $u(x,r)$ is the Hellinger scoring function: $-\hellinger(\bysinfer(\dataobs), r)$, and $GS$ is the global sensitivity of the scoring function, defined as:
\begin{equation*}
GS = 
\max_{\dataobs' \in \datauni^n: \\ \adj{\dataobs}{\dataobs'}}\max_{\{r\in \mathcal{R}\}}
|\hellinger(\bysinfer(\dataobs), r) - \hellinger(\bysinfer(\dataobs'), r)|
\end{equation*}
%
  \begin{algorithm}
  \caption{$\expmech$ in Beta-binomial model}
  \label{mech:expmech}
  \begin{algorithmic}
  \INPUT observed data set $\dataobs \in \mathcal{X}^n$, inference algorithm: $\bysinfer$, 
  prior: $\betad(\alpha, \beta)$, likelihood: $\binomial{n}{\theta}$, $\epsilon$
  \STATE \quad {\bf let} $\betad(\alpha', \beta') = \bysinfer(\binomial{n}{\theta}, \betad(\alpha, \beta), \dataobs)$. $GS$ be the global sensitivity of scoring function.  
  \STATE \quad {\bf Initialize} a vector $Pr = (0, \dots, 0 )$ of $|\candidateset|$ length.
  \STATE \quad {\bf For} $r$ in $\candidateset$:
  \STATE \quad \quad calculate the outputting probability of $r$.
  \STATE \quad \quad $Pr[r] = 
  {\exp(\frac{\epsilon\cdot \ux{r}}{2 GS})}
/ {\Sigma_{r' \in \mathcal{R}}\exp(\frac{\epsilon\cdot u(\dataobs, r')}{2\cdot GS})}$
\STATE \quad sample $z$ from $\candidateset$ by probability distribution $Pr$.
  \STATE {\bf return} $z$
  \end{algorithmic}
  \end{algorithm}

>>>>

\section{Private Mechanisms}
\label{sec_mechs}
In this section we will describe and analyse a series of mechanisms.
Section \ref{sec_ell1_dim} presents a first mechanism, $\lapmech$,
based on the Laplace primitive where the sensitivity is measured w.r.t. to $\ell_1$ norm.
Section \ref{sec_ell1_hist}, shows that $\lapmech$'s accuracy can be improved
by performing a better analysis on the sensitivity of the inference process.
Section \ref{sec_ehd}, shows an instance of the exponential mechanism, $\expmech$, where the scoring
function is based on the Hellinger distance, we will see how this mechanisms performs
poorly due to the very high global sensitivity of the inference process w.r.t to this metric.
Section \ref{sec_ehdl} introduces, $\lexpmech$, a variation of $\expmech$ which uses a lower bound
on the global sensitivity called local sensitivity.
Unfortunately, despite improving on accuracy $\lexpmech$ is non-private.
Finally, building on $\lexpmech$, Section \ref{sec_ehds} introduces $\hexpmech$:
which, based on the notion of \emph{smooth} sensitivity, achieves $\epsilon$-differential privacy and
has good accuracy in specific cases.
In presenting these mechanisms we will denote by $\inferd{\valpha}{\vtheta}{\dataobs}$ the posterior  $\dirichlet{\valpha'}$
obtained starting from the prior $\dirichlet{\valpha}$, likelihood $\multinomial{|\dataobs|}{\vtheta}$, and observed data $\dataobs$.
Two different upper bounds to the value
\begin{align}
  \label{eqn:eq1}
  \max_{\substack{\adj{\dataobs}{\dataobs'},\\
      \dirichlet{\valpha}=\inferd{\valpha}{\vtheta}{\dataobs},\\
      \dirichlet{\valpha'}=\inferd{\valpha}{\vtheta}{\dataobs'}}}||\valpha-\valpha'||_1 
\end{align}
will be used to calibrate the noise in the mechanisms based on the Laplacian primitive.

\subsection{Calibrating noise w.r.t. $\ell_1$ norm ($\lapmech$)}
\label{sec_ell1_dim}
We start off by presenting $\lapmech$, in Algorithm \ref{lapmech}, based on the Laplace primitive.
The idea of $\lapmech$ is to add Laplacian noise to the differences of the respective numeric parameters of the prior
and posterior distribution. 
\begin{algorithm}
  \caption{$\lapmech$}
  \label{lapmech}
  \begin{algorithmic}
  \INPUT $\dataobs\in\mathcal{X}^n$, $\dirichlet{\valpha}$
  \STATE \quad {\bf let} $\dirichlet{\valpha'} = \inferd{\valpha}{\vtheta}{\dataobs}$
  \STATE \quad {\bf Initialize} a vector $\tilde{\valpha} = (0, \dots, 0 )\in \mathbb{N}^{|\mathcal{X}|}$ 
  \STATE \quad {\bf For} $i = 1 \dots |\mathcal{X}| - 1$:
  \STATE \quad \quad  {\bf let} $\eta \sim \lap{0}{\frac{|\mathcal{X}|}{\epsilon}}$
  \STATE \quad \quad  $\tilde{\alpha_i}=\alpha_i + \lfloor{(\alpha_i' - \alpha_i) + \eta}\rfloor^n_0$ 
  % \RETURN $a_{k+1}$.
  \STATE \quad $\tilde{\alpha_{|\mathcal{X}|}} = \alpha_{|\mathcal{X}|} + \lfloor n - \sum_{i = 1}^{|\mathcal{X}|-1}\lfloor{(\alpha_i' - \alpha_i) + \eta_i}\rfloor^n_0 \rfloor^n_0$
  \STATE {\bf return} $\tilde{\valpha}$
  \end{algorithmic}
\end{algorithm}
\begin{lem}
  Algorithm \ref{lapmech} is $\epsilon$-d.p.
  \begin{proof} By noticing that $\mathcal{X}-1$ is an upper bound to \ref{eqn:eq1}. \end{proof}
\end{lem}

\subsection{Improving the sensitivity ($\ilapmech$)}
\label{sec_ell1_hist}
By seeing the parameters of the distributions as histograms, we can give a tighter upper bound to \ref{eqn:eq1}.
Indeed  $\lapmech$ can be reduced to a constant. Algorithm $\ilapmech$ in Algorithm \ref{mech:ilapmech} takes advantage of this fact. 
Indeed, when $\mathcal{X}=2$, \ref{eqn:eq1} is upper bounded by 1, while when $\mathcal{X}=2$ \ref{eqn:eq1} is upper bounded by 2.
In Algorithm \ref{mech:ilapmech} the algorithm $\ilapmech$ is presented.
\begin{algorithm}
  \caption{$\ilapmech$ in Dirichlet-Multinomial model}
  \label{mech:ilapmech}
  \begin{algorithmic}
  \INPUT observed data set $\dataobs \in \mathcal{X}^n$, inference algorithm: $\bysinfer$, 
  prior: $\dirichlet{\valpha}$, likelihood: $\multinomial{n}{\vtheta}$, $\epsilon$
  \STATE \quad {\bf let} $\dirichlet{\valpha'} = \bysinfer(\multinomial{n}{\vtheta}, \dirichlet{\valpha}, \dataobs)$, $k = |\mathcal{X}|$.
  \STATE \quad {\bf Initialize} a vector $\tilde{\valpha} = (0, \dots, 0 )$ of $k$ dimensions .
  \STATE \quad {\bf For} $i = 1 \dots (k - 1)$:
  \STATE \quad \quad  {\bf let} $\eta \sim \lap{0}{\frac{2}{\epsilon}}$.
  \STATE \quad \quad  $\tilde{\alpha_i}=\alpha_i + \lfloor{(\alpha_i' - \alpha_i) + \eta}\rfloor^n_0$ 
  % \RETURN $a_{k+1}$.
  \STATE \quad $\tilde{\alpha_k} = \alpha_k + \lfloor n - \sum_{i = 1}^{k-1}\lfloor{(\alpha_i' - \alpha_i) + \eta_i}\rfloor^n_0 \rfloor^n_0$
  \STATE {\bf return} $\dirichlet{\tilde{\valpha}}$
  \end{algorithmic}
  \end{algorithm}

\subsection{A new family of mechanisms}
\label{sec_exp}
In this section, we explore  mechanism which we denote by $\expmech$. $\expmech$ is an instance
of the exponential mechanism \cite{talwar} where the scoring function is based on a metric over
probability distributions.
The idea is that $\expmech$ will sample and output a distribution with higher probability when
it is closer to the real posterior; importantly,  \emph{closer} is measured using a metric over probability
distributions as opposed to a metric over the parameters that represent it. In this work we will
use the Hellinger distance, but this approach could be extended to other $f$-divergences. We will
start by describing how the mechanism works in the case of the Beta-Binomial model noticing that
the extension to the Dirichelet-Multinomial model is straightforward.
Given a prior distribution $\betad(\alpha, \beta)$ and an observed data set: $\dataobs\in\{0,1\}^n$,
we define the follwing set  where the mechanism will sample from:
\[
  \candidateset=\{\bysinfer(\binomial{n}{\theta}, \betad(\alpha, \beta),\dataobs)\mid \dataobs \in \{0,1\}^n\}.
\]
Where $\bysinfer(\binomial{n}{\theta}, \bprior,\dataobs)$ denotes the posterior beta distribution inferred starting from the parameters in input.
Notice that the above set reduces to:
\[
  \{\betad(\alpha',\beta')\mid \alpha'=\alpha + \Delta\alpha, \beta'=\beta + n-\Delta\alpha\}
\]
We use $\bysinfer(\dataobs)$ to denote the Bayesian inference process on the observed data $\dataobs$
when the other two parameters, that is prior and likelihood, are fixed.

\subsubsection{A first approach: the $\expmech$ mechanism}
\label{sec_ehd}
$\expmech(x,u,\candidateset)$  outputs an element $r\in\candidateset$ with  probability: % proportional to $\exp(\frac{\epsilon u(x,r)}{2 GS})$.
\[
%\underset{z \thicksim \expmech(x,u,\candidateset) }{\Pr}[z=r] = \frac
\frac
  {\exp(\frac{\epsilon\cdot \ux{r}}{2 GS})}
{\Sigma_{r' \in \mathcal{R}}\exp(\frac{\epsilon\cdot \ux{r'}}{2\cdot GS})},
\]
where $u(x,r)$ is the Hellinger scoring function: $-\hellinger(\bysinfer(\dataobs), r)$, and $GS$ is the global sensitivity of the scoring function, defined as:
\begin{equation*}
GS = 
\max_{\dataobs' \in \datauni^n: \\ \adj{\dataobs}{\dataobs'}}\max_{\{r\in \mathcal{R}\}}
|\hellinger(\bysinfer(\dataobs), r) - \hellinger(\bysinfer(\dataobs'), r)|
\end{equation*}
%
  \begin{algorithm}
  \caption{$\expmech$ in Beta-binomial model}
  \label{mech:expmech}
  \begin{algorithmic}
  \INPUT observed data set $\dataobs \in \mathcal{X}^n$, inference algorithm: $\bysinfer$, 
  prior: $\betad(\alpha, \beta)$, likelihood: $\binomial{n}{\theta}$, $\epsilon$
  \STATE \quad {\bf let} $\betad(\alpha', \beta') = \bysinfer(\binomial{n}{\theta}, \betad(\alpha, \beta), \dataobs)$. $GS$ be the global sensitivity of scoring function.  
  \STATE \quad {\bf Initialize} a vector $Pr = (0, \dots, 0 )$ of $|\candidateset|$ length.
  \STATE \quad {\bf For} $r$ in $\candidateset$:
  \STATE \quad \quad calculate the outputting probability of $r$.
  \STATE \quad \quad $Pr[r] = 
  {\exp(\frac{\epsilon\cdot \ux{r}}{2 GS})}
/ {\Sigma_{r' \in \mathcal{R}}\exp(\frac{\epsilon\cdot u(\dataobs, r')}{2\cdot GS})}$
\STATE \quad sample $z$ from $\candidateset$ by probability distribution $Pr$.
  \STATE {\bf return} $z$
  \end{algorithmic}
  \end{algorithm}
